---
title: "scratch for blog"
author: "dee"
date: "6/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# TO-DO



* make note that it is possible to create definition by code - outside of scope of this. **add a link??**






#  vvv TO DELETEMORE DETAIL vvvv #







To go over these parameters in a little more detail:

* `collection` The IPUMS data collection to query, abbreviated and lower-case. Currently only USA and CPS are supported, and should be specified in **lowercase abbreviations:** 
  + "usa", "cps"

* `data_dir` Where to download your data (will be created if it does not exist). We recommend storing data, dictionaries, and extract definitions in a sub-folder of your R project. The default will create a "Data." folder within. 
* If you want to store your files at the **top-level** of the project directory use: 
  + `data_dir <- file.path("")`.
* If you want to store your files **outside** of the project directory use:
  + `data_dir <- file.path("..","Data")` to store in a sibling-directory to the project directory.
  + The `".."` goes "up" one level within a folder system.

`descriptive_name` IPUMS provides numerical IDs for each data extract by default (eg, `usa_000001.dat.gz, usa_000001.xml`), however these are specific to individual users and can be confusing to keep track of. We recommend users relabel their extracts using a project- or analysis-specific descriptive name, eg: "prcs_migration_ex". The script will automatically apply the same `descriptive_name` to the `.json, .dat.gz, .xml`, as well as a `.csv` file used for checking extract status.

**extract_num** The last step is to indicate an extract definition. If you've already created an extract online, via the Data Cart GUI, or someone has shared a .json extract definition with you, follow one of the following options. 
* To work with a **past extract**
  + Take note of the **Extract Number** on your [Download or Revise Data page](https://usa.ipums.org/usa-action/data_requests/download).
  + Modify `extract_num <- 42`
  + **screenshot of where this is/looks like**

* To work with your **most recent extract**.
  + Leave `extract_num <- ""` as is
  
* If you **already have a .json extract definition**
  + Manually add it to the `data_dir` folder

The microdata API **does** allow users to build extracts 'by hand' using the `define_extract_micro()` function. Users can specify arugments for this in the parameters of the `RRR` template. If you would like to do so, leave `extract_num <-""` as in, and modify `extract_definition` as needed. It should be noted that big extracts could quickly become unwieldy to type out, and this approach require users to know the IPUMS mnemonics in order to specify variables, samples, etc. In practice, it might be easier to build your extract online, until you become more familiar with the shorthand. Hard coding an extract definition takes just 4 arguments to `define_extract_mictro()`.

And within that
**passing dinner gif?**

The template makes as few assumptions as possible, yet automates as many steps as possible. Upon opening the template, users will be prompted for 4 key parameters. With these in place,the script will check on, submit, and download data straight to your specified directory. Once compiled, only the script itself and the `.json` file are needed to share/reproduce the work in this script - or you could provide the interactive .html report. 

**maybe use a screenshot instead**

**2, 1 empty, 1 filled**

```{r, echo = FALSE}

## screenshots

```

## Build An Extract Definition

The first step to any analysis using IPUMS data is to decide on a a sample/variable set. This can now be done **entirely through code** with the `define_extract_micro()` function. By specifying an IPUMS `collection`, `samples`, and `variables`, users can create an `ipums_extract` object, and `save_extract_as_json()` or `submit_extract()` to the IPUMS servers.

```{r, echo = TRUE, eval = FALSE}

my_new_extract <- define_extract_micro(collection = "usa",
                                    samples = "us2019b",
                                    variables = c("AGE", "SEX", "INCTOT", "EDUC"),
                                    description = "A simple extract"
)

my_new_extract %>% save_extract_as_json("new_extract.json")

my_new_extract %>% submit_extract()

```

At this stage, you have everything you need to share your IPUMS data. Just email a colleague the .json file, so they can `define_extract_from_json()` and `submit_extrat()` to get their own version of the data. Note, that `collection = "usa"` must be specified again when reading from a .json file. 

```{r. echo = TRUE, eval = FALSE}

extract_from_friend <- define_extract_from_json("my_new_extract.json", collection = "usa")

extract_from_friend %>% submit_extract()

```



creating a .json extract definition, submitting, checking, and downloading datathat allow users to interact with the microdata API, and accomplish all these steps. these brandwe've bundled several new functions together into a new [.Rmd template]() to streamline the steps of creating truly reproducible research with IPUMS data. Check our  [interactive report]() to see the **new API template** in action. This is accomplished by a series of helper functionsIn fact, the single most handy feature - in this users opinion - is being able to download **BOTH** data **AND** data dictionary 



Build initial extract definition online (as before)


The microdata API allows users to build extracts totally 'by hand'. This requires users to know the IPUMS mnemonics in order to specify variables. If you know what you need, fill in/add relevant arguments to `extract_definition` in the section below. Otherwise, if you have  previously defined an extract using the online system, simply take note of the **extract number** and enter it below. If you know your **most recent** extract is the one you'd like to use or you are building the extract **'by hand'**, leave `extract_num <- ""` as is. 

If you already have a `.JSON` extract definition from another source, just add it to the `data_dir` folder and this script will read and submit that extract. 





First we'll walk through the steps of setting up reproducible research project, then we'll talk about how these get implemented in the **.Rmd for Reproducible Research** in action, as we **create .json extract definitions**, **submit, (check on), and download data**, to create a [portable, reproducible, literately-programed report]()!!

*if derek's post DOES NOT cover basic use of functions, include that here, otherwise just focus on the template and how they fit together*


But you realize you can just send them instructions to recreate your extract:

Just a few lists:give them the instructions to rebuild your extract, it's so easy! just a couple...

[ipums project]()
[get data]()

sreenshots...

[submit extract]()
[downloads page]() 

Oh and remember:

[download page with save as directions]()


At this point, some of you may be asking (or screaming): "isn't there an easier way to share IPUMS data???"

Have you ever tried to share your items data and then remembered oh wait I can't share my data. The items micro data API and items are packaged together let you submit a data extract compile an extract request save it as a JSON file submit that request to the R servers check the status and see if it's available and ready for download and finally download those files both data and the metadata in the DDI directly to whichever project directory you want we've combined many of these steps together into a new R markdown template available in the FM's R package together these steps make it really easy to reproduce and share your research with colleagues

variable viewer of value variance