---
title: "Graphing migration in Puerto Rico 2015-2019 with IPUMS USA PRCS data"
author: "Your Name Here"
date: '2022-04-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
suppressPackageStartupMessages(suppressWarnings({
  library(ipumsr)
  library(tidyverse)
}))

data_path <- file.path("Data")  
```

```{r load-data, echo=FALSE}
if (!file.exists("Data/prcs_migration_extract.xml")) {
  # Load extract definition from JSON
  prcs_migration_extract <- define_extract_from_json(
    "Data/prcs_migration_extract.json",
    "usa"
  )
  # Submit, wait for, and download extract
  ddi_filename <- submit_extract(prcs_migration_extract) %>% 
    wait_for_extract() %>% 
    download_extract(download_dir = data_path) %>% 
    basename()
  # Infer data file name from DDI file name
  data_filename <- str_replace(ddi_filename, "\\.xml$", ".dat.gz")
  # Standardize DDI and data file names 
  file.rename(file.path(data_path, ddi_filename),
              file.path(data_path, "prcs_migration_extract.xml"))
  file.rename(file.path(data_path, data_filename),
              file.path(data_path,"prcs_migration_extract.dat.gz"))
}
ddi <- read_ipums_ddi("Data/prcs_migration_extract.xml")
data <- read_ipums_micro(ddi, data_file = "Data/prcs_migration_extract.dat.gz")
```

```{r prep-data}
# Prep education variable
college_regex <- "^[123] year(s)? of college$"
data <- data %>% 
  mutate(
    EDUCD3 = EDUCD %>%
      lbl_collapse(~.val %/% 10) %>% 
      lbl_relabel(
        lbl(2, "Less than High School") ~.val > 0 & .val < 6,
        lbl(3, "High school") ~.lbl == "Grade 12", #<<
        lbl(4, "Some college") ~str_detect(.lbl, college_regex), #<<
        lbl(5, "College or more") ~.val %in% c(10, 11)
      ) %>%
      as_factor()
  )

# Prep income variable
value_to_quintile <- function(x) {
  cut_points <- quantile(x, probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE)
  cut(
    x, 
    breaks = c(-Inf, cut_points, Inf), 
    labels = c("Lowest", "Lower Middle", "Middle", "Upper Middle", "Highest"),
    ordered_result = TRUE
  )
}

hhincome_quintiles <- data %>% 
  filter(PERNUM == 1 & HHINCOME != 9999999) %>% 
  group_by(YEAR) %>% 
  mutate(hhincome_quintile = value_to_quintile(HHINCOME)) %>% 
  ungroup() %>% 
  select(YEAR, SERIAL, hhincome_quintile)

data <- data %>% 
  left_join(hhincome_quintiles, by = c("YEAR", "SERIAL"))

# Prep migration variable
data <- data %>% 
  mutate(
    moved_in_last_year = case_when(
      MIGRATE1 %in% c(0, 9) ~ NA, 
      MIGRATE1 == 1 ~ FALSE, 
      MIGRATE1 %in% 2:4 ~ TRUE
    )
  )


# Prep age variable
age_to_age_group <- function(x) {
  cut_points <- c(9, 17, 34, 64)
  cut(
    x,
    breaks = c(-Inf, cut_points, Inf),
    labels = c("0-9", "10-17", "18-34", "35-64", "65+"),
    ordered_result = TRUE
  )
}

data <- data %>% 
  mutate(age_group = age_to_age_group(AGE))
```

# Migration 2015-2019 {.tabset}

The percentage of people who had moved in the last year increased between 2017 
and 2018 from about 6% to over 8% among all persons in Puerto Rico, but the 
magnitude of this trend varies by education, household income, and age.

Note: These graphs show trends in point estimates from sample data, without 
displaying estimates of sampling error. Differences over time or across groups 
may not be statistically significant. To calculate confidence intervals for 
point estimates, follow the 
[IPUMS USA instructions for using replicate weights](https://usa.ipums.org/usa/repwt.shtml).

## Overall

```{r migration-graph-1, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  group_by(YEAR) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT)
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```

## By education

```{r migration-graph-2, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  filter(AGE >= 25) %>%
  group_by(YEAR, EDUCD3) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~EDUCD3) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      subtitle = "Among persons age 25 and older",
      x = NULL,
      y = "%"
    )
```

## By household income

```{r migration-graph-3, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  filter(!is.na(hhincome_quintile)) %>% 
  group_by(YEAR, hhincome_quintile) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~hhincome_quintile) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```


## By age

```{r migration-graph-4, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  group_by(YEAR, age_group) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~age_group) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```


## Setup Code

The code below defines the steps of checking for IPUMS data within within the specified `Data` subfolder. If the appropriate `.dat.gz, .xml` files **do not** exist, it triggers code to:
* CREATE a [data extract]() from the included JSON definition. 
* SUBMIT the extract to IPUMS data servers.
* WAIT and periodically check on status of data extract, when ready...
* DOWNLOAD `data` and `data dictionary` to specific directory.
* RENAME files from `usa_000030.xml, usa_000030.dat.gz` to an explicit file name.
* READ IN both `data` and `data dictionary` to current session.

If the file is re-run at a later date, the `.dat.gz` and `.xml` data files will be detected skipping ahead to simply loading these files into R as `data` and `ddi`, respectively. Starting your 
From here, we can 

```{r setup_code, eval = FALSE}


  library(ipumsr)
  library(tidyverse)


## extract definition

## IF the 
json_filename <- "prcs_migration_extract.json"
data_rename <- "prcs_migration_extract.dat.gz"
ddi_rename <- "prcs_migtation_extract.xml" 

collection <- "usa" ## c("usa", "cps")
## maybe include a WAIT_NOW parameter

data_path <- file.path("Data")  ## Specify a sub-directory or filepath

if (!(file.exists(file.path(data_path, data_filename)) & 
      file.exists(file.path(data_path, dict_filename)))) {
  # Load extract definition from JSON
  extract_definition <- define_extract_from_json(
    file.path(data_path, dict_filename),
    collection
  )
  # Submit, wait for, and download extract
  ddi_filename <- submit_extract(extract_definition) %>% 
    wait_for_extract() %>% 
    download_extract(download_dir = data_path) %>% 
    basename()
  # Infer data file name from DDI file name
  data_filename <- str_replace(ddi_filename, "\\.xml$", ".dat.gz")
  # Standardize DDI and data file names 
  file.rename(file.path(data_path, ddi_filename),
              file.path(data_path, "prcs_migration_extract.xml"))
  file.rename(file.path(data_path, data_filename),
              file.path(data_path,"prcs_migration_extract.dat.gz"))
}

ddi <- read_ipums_ddi("Data/prcs_migration_extract.xml")
data <- read_ipums_micro(ddi, data_file = "Data/prcs_migration_extract.dat.gz")

```