---
title: "Reproducible Research"
author: 
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
---

# {.tabset}

## Load Packages

```{r load-packages}
suppressPackageStartupMessages({
  library(ipumsr)
  library(tidyverse)
})
```


## Define File Paths

```{r define-file-paths}
extract_definition_path <- "data/prcs_migration_analysis.json"
data_path <- "data/prcs_migration_analysis.dat.gz"
ddi_path <- "data/prcs_migration_analysis.xml"
```


## Load your IPUMS Data

This analysis of IPUMS data is designed to be shared, and thus does not assume 
that you have already downloaded the data used in the analysis. The code below 
checks whether the data are already downloaded, and if they aren't, it submits 
a new IPUMS extract request according to the specifications in the included 
extract definition JSON file.

```{r check-for-data, class.source = "fold-hide"}
# Define path to "waiting_for_extract" flag file
data_dir <- dirname(extract_definition_path)
waiting_for_extract_path <- file.path(data_dir, "waiting_for_extract.txt")

# Ensure the JSON extract definition is present
json_file_exists <- file.exists(extract_definition_path)
if (!json_file_exists) {
  stop(
    "File '", extract_definition_path, "' not found; make sure that ",
    "`extract_definition_path` is the path to the .json extract definition ", 
    "file.", call. = FALSE
  )
}

# Are the data downloaded, or are we waiting for an extract?
data_not_downloaded <- !file.exists(data_path)
data_downloaded <- file.exists(data_path)
waiting_for_extract <- file.exists(waiting_for_extract_path)

# Ensure that IPUMS_API_KEY environment variable is defined
ipums_api_key_undefined <- Sys.getenv("IPUMS_API_KEY") == ""
if (data_not_downloaded & ipums_api_key_undefined) {
  stop(
    "Environment variable 'IPUMS_API_KEY' is undefined. Make sure you've ", 
    "followed the instructions under 'Setting up your API key' in the ", 
    "ipumsr API vignette (`vignette(\"ipums-api\", package = \"ipumsr\")`) ", 
    "before running this script.", call. = FALSE
  )
}

# If not yet waiting for extract, create and submit a new extract and create 
#   the "waiting_for_extract" flag file
if (data_not_downloaded & !waiting_for_extract) {
  extract_definition <- define_extract_from_json(extract_definition_path)
  submitted_extract <- submit_extract(extract_definition)
  writeLines(
    paste0(submitted_extract$collection, ":", submitted_extract$number),
    con = waiting_for_extract_path
  )
  waiting_for_extract <- TRUE
}

# If waiting for an extract, read extract ID from flag file and check the status
if (data_not_downloaded & waiting_for_extract) {
  extract_id <- readLines(waiting_for_extract_path)
  extract_info <- get_extract_info(extract_id)
  extract_is_ready <- is_extract_ready(extract_info)
  extract_is_stale <- !extract_is_ready & extract_info$status == "completed"
  if (extract_is_stale) {
    stop(
      paste0(
        "The data files for ", extract_info$collection, " extract number ", 
        extract_info$number, " have been removed from IPUMS servers. ",
        "Please delete the file '", waiting_for_extract_path, "' and re-run ", 
        "the template."
      ),
      call. = FALSE
    )
  }
  
  # If the extract is ready, download files and rename to match the JSON file, 
  #   then delete the waiting_for_extract flag file
  if (extract_is_ready) {
    orig_ddi_path <- download_extract(extract_info, download_dir = data_dir)
    orig_data_path <- gsub("\\.xml$", ".dat.gz", orig_ddi_path)
    ddi_file_successfully_renamed <- file.rename(orig_ddi_path, ddi_path)
    data_file_successfully_renamed <- file.rename(orig_data_path, data_path)
    if (!ddi_file_successfully_renamed || !data_file_successfully_renamed) {
      stop(
        "Problem renaming DDI and/or data file; please report bug at ", 
        "https://github.com/ipums/ipumsr/issues, including a copy of this ", 
        "file if possible.", call. = FALSE
      )
    }
    data_downloaded <- TRUE
    waiting_file_successfully_removed <- file.remove(waiting_for_extract_path)
    if (!waiting_file_successfully_removed) {
      stop(
        "Unable to remove 'waiting_for_extract.txt'; please report bug at ", 
        "https://github.com/ipums/ipumsr/issues, including a copy of this ", 
        "file if possible.", call. = FALSE
      )
    }
  } else { # If extract is not ready, stop execution
    stop(
      "NOT AN ERROR: ", extract_info$collection, " extract number ", 
      extract_info$number, " is not yet ready to download. Try ", 
      "re-running again later.", call. = FALSE
    )
  }
}
```

```{r load-data, eval = data_downloaded}
ddi <- read_ipums_ddi(ddi_path)
data <- read_ipums_micro(ddi, data_file = data_path)
```


## Analysis: Migration in Puerto Rico 2015-2019 {.tabset}

The percentage of people who had moved in the last year increased between 2017 
and 2018 from about 6% to over 8% among all persons in Puerto Rico, but the 
magnitude of this trend varies by education, household income, and age.

Note: These graphs show trends in point estimates from sample data, without 
displaying estimates of sampling error. Differences over time or across groups 
may not be statistically significant. To calculate confidence intervals for 
point estimates, follow the 
[IPUMS USA instructions for using replicate weights](https://usa.ipums.org/usa/repwt.shtml).

```{r prep-data}
# Prep education variable
college_regex <- "^[123] year(s)? of college$"
data <- data %>% 
  mutate(
    EDUCD3 = EDUCD %>%
      lbl_collapse(~.val %/% 10) %>% 
      lbl_relabel(
        lbl(2, "Less than High School") ~.val > 0 & .val < 6,
        lbl(3, "High school") ~.lbl == "Grade 12", #<<
        lbl(4, "Some college") ~str_detect(.lbl, college_regex), #<<
        lbl(5, "College or more") ~.val %in% c(10, 11)
      ) %>%
      as_factor()
  )

# Prep income variable
value_to_quintile <- function(x) {
  cut_points <- quantile(x, probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE)
  cut(
    x, 
    breaks = c(-Inf, cut_points, Inf), 
    labels = c("Lowest", "Lower Middle", "Middle", "Upper Middle", "Highest"),
    ordered_result = TRUE
  )
}

hhincome_quintiles <- data %>% 
  filter(PERNUM == 1 & HHINCOME != 9999999) %>% 
  group_by(YEAR) %>% 
  mutate(hhincome_quintile = value_to_quintile(HHINCOME)) %>% 
  ungroup() %>% 
  select(YEAR, SERIAL, hhincome_quintile)

data <- data %>% 
  left_join(hhincome_quintiles, by = c("YEAR", "SERIAL"))

# Prep migration variable
data <- data %>% 
  mutate(
    moved_in_last_year = case_when(
      MIGRATE1 %in% c(0, 9) ~ NA, 
      MIGRATE1 == 1 ~ FALSE, 
      MIGRATE1 %in% 2:4 ~ TRUE
    )
  )


# Prep age variable
age_to_age_group <- function(x) {
  cut_points <- c(9, 17, 34, 64)
  cut(
    x,
    breaks = c(-Inf, cut_points, Inf),
    labels = c("0-9", "10-17", "18-34", "35-64", "65+"),
    ordered_result = TRUE
  )
}

data <- data %>% 
  mutate(age_group = age_to_age_group(AGE))
```

### Overall

```{r migration-graph-1, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  group_by(YEAR) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT)
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```

### By education

```{r migration-graph-2, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  filter(AGE >= 25) %>%
  group_by(YEAR, EDUCD3) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~EDUCD3) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      subtitle = "Among persons age 25 and older",
      x = NULL,
      y = "%"
    )
```

### By household income

```{r migration-graph-3, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  filter(!is.na(hhincome_quintile)) %>% 
  group_by(YEAR, hhincome_quintile) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~hhincome_quintile) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```


### By age

```{r migration-graph-4, dpi=300, fig.height=5, fig.width=8, echo = FALSE}
data %>% 
  filter(!is.na(moved_in_last_year)) %>% 
  group_by(YEAR, age_group) %>% 
  summarize(
    pct_moved = 100 * sum(PERWT[moved_in_last_year]) / sum(PERWT),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = YEAR, y = pct_moved)) +
    geom_line() +
    facet_wrap(~age_group) +
    labs(
      title = "Percentage of people who moved in the past year, 2015-2019",
      x = NULL,
      y = "%"
    )
```
